Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-09-19 Manuscript_Nater_etal.qmd
Fixed a few mispellings and made a few minor comments.

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structured citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in reproducible and semi-automated ways that increase their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reasons for the long-term decline in abundance remain unresolved, but the willow ptarmigan is considered a sentinel species that is sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey for resident predators, such as the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020]. Population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycles through shared predators [@hagen1952; @bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988]. Across their distributional range, willow ptarmigan are an iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

![Graphical representation of the ptarmigan life cycle with two age classes and the data sources, as well as joint distributions, included in the integrated distance sampling model. Indices $t$ denote year while $x$ represent the added dimension for area (corresponding to 41 reporting districts). Juvenile~x,t~ = juveniles in area $x$ year $t$. Adult~x,t~ = adults in area $x$ in year $t$. $R_{x,t}$ = recruitment rate in area $x$ in year $t$. $S_{x,t}$ = survival probability from year $t$ to $t+1$ in area $x$. Note that the additional site (=transect) dimension, $j$, is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}

@@

The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model). For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for one study area (the Lierne area).

@@

For presentation and interpretation of results, we calculated additional key demographic parameters from the model posteriors. First, we calculated area- and year-specific average population densities, $meanDens_{x, t}$, by averaging $D_{juv, x, j, t} + D_{ad, x, j, t}$ over all transect lines $j$ within area $x$ in year $t$. We then proceeded to derive area- and year-specific realized population growth rates as $meanDens_{x, t+1} / meanDens_{x, t}$. Additionally, we calculated generation time for each area using two different approaches: per-generation population growth rate [@caswell2000] and elasticity of asymptotic growth rate with respect to fecundity [@brooks2001].

@@

Based on estimates of population growth rates and vital rates, we also calculated generation time as both per-generation population growth rate ($R0$) and inverse of fecundity elasticity ($1/elas_F$). The two approaches yielded very similar estimates (median correlation coefficient = `r medianCorr`) ranging from `r GT_min_print` to `r GT_max_print` years across areas (SF "GenerationTime_Latitude.pdf"). Spatial patterns in generation time were consistent with those for survival and recruitment, with the highest values occurring in the North and in the mountaineous regions in central Norway (SFs "GenerationTime_R0_Map.pdf" and "GenerationTime_elasF_Map.pdf").

@@

The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`). Nonetheless, spatial variation in the rodent effect was non-negligible (random effect SD = `r betaR.sd.spatial$post.sum`). This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf"). The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf"). Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival, lower recruitment, and longer generation times) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring that employs common sampling protocol, training programmes, and data processing pipelines. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on fine-scale spatio-temporal variation in demography and population dynamics at across a large area. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and rugged terrain, which may hamper detectability. Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g. @bowler2020, see also next section]. Our modelling framework can be easily adapted for studying the impact of these and other variables on detectability (see below). Additionally, our results could motivate taking a closer look at monitoring challenges and potential improvements in the northern part of the country in particular. Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.

>>>>

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structured citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal <!--# we don't hyphenate this in the title. Not sure which is best but we should be consistent -->variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in reproducible and semi-automated ways that increase their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

The willow ptarmigan is a tetraonid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reasons for the long-term decline in abundance remain unresolved, but the willow ptarmigan is considered a sentinel species that is sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey for resident predators, such as the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020]. Population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycles through shared predators [@hagen1952; @bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988]. Across their distributional range, willow ptarmigan are an iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

![Graphical representation of the willow ptarmigan life cycle with two age classes and the data sources, as well as joint distributions, included in the integrated distance sampling model. Indices $t$ denote year while $x$ represent the added dimension for area (corresponding to 41 reporting districts). Juvenile~x,t~ = juveniles in area $x$ year $t$. Adult~x,t~ = adults in area $x$ in year $t$. $R_{x,t}$ = recruitment rate in area $x$ in year $t$. $S_{x,t}$ = survival probability from year $t$ to $t+1$ in area $x$. Note that the additional site (=transect) dimension, $j$, is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}

@@

The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios; see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model). For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for one study area (the Lierne area).

@@

For presentation and interpretation of results, we calculated additional key demographic parameters from the model posteriors. First, we calculated area- and year-specific average population densities, $meanDens_{x, t}$, by averaging $D_{juv, x, j, t} + D_{ad, x, j, t}$ over all transect lines $j$ within area $x$ in year $t$. We then proceeded to derive area- and year-specific realized population growth rates as $meanDens_{x, t+1} / meanDens_{x, t}$. Additionally, we calculated generation time for each area using two different approaches: per-generation population growth rate [@caswell2000] and elasticity of asymptotic growth rate with respect to fecundity [@brooks2001]. <!--# Do we want to say recruitement instead of fecundity? -->

@@

Based on estimates of population growth rates and vital rates, we also calculated generation time as both per-generation population growth rate ($R0$) and inverse of fecundity elasticity ($1/elas_F$). The two approaches yielded very similar estimates (median correlation coefficient = `r medianCorr`) ranging from `r GT_min_print` to `r GT_max_print` years across areas (SF "GenerationTime_Latitude.pdf"). Spatial patterns in generation time were consistent with those for survival and recruitment, with the highest values occurring in the North and in the mountainous regions in central Norway (SFs "GenerationTime_R0_Map.pdf" and "GenerationTime_elasF_Map.pdf").

@@

The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`). Nonetheless, spatial variation in the rodent effect was non-negligible (random effect SD = `r betaR.sd.spatial$post.sum`). This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf"). The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountainous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf"). Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. <!--# Do we want to start a new paragraph after this sentence? -->

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival, lower recruitment, and longer generation times) were more common in the northern and mountainous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential within coordinating structured monitoring that employs common sampling protocol, training programmes, and data processing pipelines. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on fine-scale spatio-temporal variation in demography and population dynamics at across a large area. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and rugged terrain, which may hamper detectability. Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g. @bowler2020, see also next section]. Our modelling framework can be easily adapted for studying the impact of these and other variables on detectability (see below). Additionally, our results could motivate taking a closer look at monitoring challenges and potential improvements in the northern part of the country in particular. Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
